---
title: "Gonadal but not hepatic transcriptomes distinguish territorial from social sockeye salmon (*Oncorhynchus nerka*)"
author:
  - name: "[[Author list to be completed]]"
    affiliations:
      - "School of Aquatic and Fishery Sciences, University of Washington, Seattle, WA, USA"
date: today
abstract: |
  Social phenotype shapes physiology across vertebrates, but in wild salmonids the
  molecular correlates of behavioural type have been studied almost exclusively in
  the brain. We used 3′-tag RNA sequencing (TagSeq) to compare gonad and liver
  transcriptomes of adult sockeye salmon classified in the field as territorial or
  social (n = 15 per phenotype, two tissues per fish). Gonadal expression differed
  markedly between phenotypes: 1,653 of 18,156 tested genes were
  differentially expressed (adjusted p < 0.05), 79% of them more highly
  expressed in territorial fish. Over-representation analysis of these genes recovered
  a coherent programme of elevated protein synthesis and energy production, led by
  ribosome and oxidative phosphorylation pathways (5 KEGG pathways and
  36 GO terms at FDR < 0.05), alongside strong induction of acute-phase
  and stress-response genes including two serum amyloid A paralogues and a small heat
  shock protein. By contrast the liver differed only slightly between
  phenotypes, with 66 differentially expressed genes of 4,076
  tested and no enriched pathway. Expression of sex-linked marker genes indicated that
  all gonad samples were ovarian. The results point to the gonad, rather than the
  liver, as the peripheral tissue in which the territorial phenotype of adult sockeye
  is physiologically expressed, and are consistent with territorial females being in
  a more active state of reproductive investment. Because behaviour was scored at
  capture in a cross-sectional design, the direction of causation between behavioural
  type and gonadal state cannot be resolved here. All analyses are fully scripted,
  version-pinned, and re-executed on an independent machine as part of a continuous
  integration check that reproduces every reported number.
keywords: [sockeye salmon, social behaviour, territoriality, TagSeq, gonad, liver, transcriptomics, oxidative phosphorylation]
bibliography: references.bib
link-citations: true
number-sections: true
execute:
  echo: false
  warning: false
  message: false
format:
  html:
    toc: true
    embed-resources: true
  docx:
    toc: false
    number-sections: true
---

```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages({ library(dplyr); library(knitr) })
## Resolve the project root (the directory holding _quarto.yml) so this document
## renders identically whether Quarto executes it from the project root or from
## manuscript/. All data paths below are relative to that root.
root <- normalizePath(getwd())
while (!file.exists(file.path(root, "_quarto.yml")) && dirname(root) != root) root <- dirname(root)
setwd(root)
knitr::opts_knit$set(root.dir = root)
source("tag-seq/code/_common.R")

fmt <- function(x) format(x, big.mark = ",", trim = TRUE)
pct <- function(a, b, d = 0) paste0(formatC(100 * a / b, format = "f", digits = d), "%")
sci <- function(p) {
  if (p >= 0.001) return(formatC(p, format = "f", digits = 3))
  e <- floor(log10(p)); m <- p / 10^e
  sprintf("%.1f &times; 10^%d^", m, e)   # HTML entity: R's sprintf mangles non-ASCII under a C locale
}

read_tissue <- function(tissue) {
  cfg <- tissue_config(tissue)
  list(
    all = read.csv(out_path(cfg, cfg$prefix, "-ALL-DEG-apeglm.csv"), stringsAsFactors = FALSE),
    sig = read.csv(out_path(cfg, cfg$prefix, "-SIG-DEG-apeglm.csv"), stringsAsFactors = FALSE),
    counts = read.csv(out_path(cfg, cfg$prefix, "-gene-counts.csv"), row.names = 1),
    go  = read.csv(file.path("tag-seq/GO_output", tissue, paste0(tissue, "-GO-all.csv")), stringsAsFactors = FALSE),
    kegg = read.csv(file.path("tag-seq/GO_output", tissue, paste0(tissue, "-KEGG-all.csv")), stringsAsFactors = FALSE)
  )
}
L <- read_tissue("liver"); G <- read_tissue("gonad")

l_tested <- nrow(L$all); l_sig <- nrow(L$sig); l_up <- sum(L$sig$log2FoldChange > 0); l_down <- l_sig - l_up
g_tested <- nrow(G$all); g_sig <- nrow(G$sig); g_up <- sum(G$sig$log2FoldChange > 0); g_down <- g_sig - g_up
l_kept <- L$counts["genes_after_filtering", "count"]; g_kept <- G$counts["genes_after_filtering", "count"]
n_genes <- L$counts["genes_before_filtering", "count"]
g_lfc1 <- sum(abs(G$sig$log2FoldChange) > 1)
g_medlfc <- median(abs(G$sig$log2FoldChange)); l_medlfc <- median(abs(L$sig$log2FoldChange))
shared <- intersect(L$sig$gene, G$sig$gene)
shared_same_dir <- sum(sign(L$sig$log2FoldChange[match(shared, L$sig$gene)]) ==
                       sign(G$sig$log2FoldChange[match(shared, G$sig$gene)]))

go_sig  <- sum(G$go$FDR  < 0.05);  k_sig  <- sum(G$kegg$FDR < 0.05)
lgo_sig <- sum(L$go$FDR  < 0.05);  lk_sig <- sum(L$kegg$FDR < 0.05)
oxphos <- G$kegg[G$kegg$category == "one00190", ]
ribo   <- G$kegg[G$kegg$category == "one03010", ]
## Enriched gonad pathways after the two leading ones, in p-value order.
k_rest <- tolower(G$kegg$name[G$kegg$FDR < 0.05 & !G$kegg$category %in% c("one00190", "one03010")])
## Best liver pathway, to report how close the liver came.
lk_top <- L$kegg[which.min(L$kegg$FDR), ]

## Variance explained by each principal component, written by notebook 05
## alongside the figure that shows it.
pv  <- read.csv("figures/source-data/pca-variance.csv")
pcv <- function(t, pc) sprintf("%.1f%%", pv$percent_variance[pv$tissue == t & pv$PC == pc])

## Library depth from the count matrices themselves.
depth <- sapply(c("liver", "gonad"), function(t) colSums(load_counts(tissue_config(t))$cts))
depth_txt <- function(t) sprintf("%.1f &plusmn; %.1f million", mean(depth[[t]]) / 1e6, sd(depth[[t]]) / 1e6)

## Direction of DE genes within the enriched KEGG pathways.
ann <- load_gene_annotation()
g2g <- setNames(ann$gene, as.character(ann$GeneID))
kl <- read.delim(file.path(PATHS$genome, "kegg_one_gene2pathway.tsv"), header = FALSE, col.names = c("gene", "pathway"))
kl$gene <- unname(g2g[sub("^one:", "", kl$gene)]); kl$pathway <- sub("^path:", "", kl$pathway)
pw_dir <- function(pw) { g <- G$sig[G$sig$gene %in% kl$gene[kl$pathway == pw], ]; c(n = nrow(g), up = sum(g$log2FoldChange > 0)) }
ox_dir <- pw_dir("one00190"); ribo_dir <- pw_dir("one03010")

## Length association (the reason the hypergeometric test is used).
len <- setNames(ann$length, ann$gene)
dec <- G$all %>% filter(gene %in% names(len)) %>% mutate(len = len[gene], de = padj < 0.05, decile = ntile(len, 10)) %>%
  group_by(decile) %>% summarise(pctDE = 100 * mean(de), .groups = "drop")
dec1 <- dec$pctDE[1]; dec_rest <- range(dec$pctDE[-1])

## Gene tables
gene_tbl <- function(sig, n = NULL, by = c("padj", "lfc")) {
  by <- match.arg(by)
  x <- if (by == "padj") sig %>% arrange(padj) else sig %>% arrange(desc(abs(log2FoldChange)))
  if (!is.null(n)) x <- head(x, n)
  x %>% transmute(Gene = gene, Description = description,
                  `log2 fold change` = formatC(log2FoldChange, format = "f", digits = 2),
                  `Adjusted p` = sapply(padj, sci))
}
```

# Introduction

Social behaviour is among the most conspicuous sources of phenotypic variation in
salmonid fishes. As juveniles in streams, individuals defend feeding territories and
form dominance hierarchies in which rank predicts growth, metabolic rate and
life-history trajectory [@Metcalfe1995; @Cutts2002; @Reid2011; @Gilmour2005]. As
adults, the same species migrate and move in aggregations whose collective
properties can improve navigation and passage past obstacles [@Berdahl2014;
@Okasaki2020], and then, on the spawning grounds, revert to intense intra-sexual
competition in which males contest access to females and females contest and defend
nest sites [@Quinn1999; @Fleming1998]. The energetic cost of these behaviours is
substantial: aggressive display and nest defence are the most expensive activities in
the daily budget of spawning Pacific salmon after simple holding [@McVeigh2007].
Whether an individual behaves territorially or remains within a social aggregation is
therefore a decision with physiological consequences, but the tissue-level molecular
correlates of that decision in wild adult salmon are largely unknown.

Transcriptomic studies of social phenotype in fish have concentrated on the brain.
Dominant and subordinate individuals differ in neural gene expression in cichlids
[@Renn2008; @Renn2016], zebrafish [@Filby2010], weakly electric fish [@Eastman2020]
and juvenile Atlantic salmon [@Roberge2008]. Fewer studies have asked how social
phenotype is reflected in peripheral tissues, yet the evidence that exists suggests
the effects can be as large as those in the brain. In Nile tilapia, dominance is
coupled to gonadotropin signalling, sex-steroid levels and testis transcription
[@Thonnes2022; @HollanderCohen2022]; in juvenile rainbow trout, four days of
subordination reorganise hepatic lipid metabolism [@Kostyniuk2018]; and in mammals,
social status remodels immune gene regulation in blood, liver and spleen
[@SnyderMackler2016; @Lee2022]. The gonad and the liver are natural targets in
salmon: the gonad because behaviour on the spawning grounds is inseparable from
reproductive state, and the liver because it integrates energy allocation and
synthesises the plasma proteins, including vitellogenin and acute-phase reactants,
that mediate the costs of competition.

Here we compare gonad and liver transcriptomes of adult sockeye salmon
(*Oncorhynchus nerka*) classified at capture as territorial or social. We used
3′-tag RNA sequencing (TagSeq), which counts one fragment per transcript and
therefore quantifies expression at a fraction of the cost of whole-transcript
RNA-seq without the transcript-length bias that complicates downstream inference
[@Meyer2011; @Lohman2016; @Moll2014]. We asked (i) whether the two behavioural
phenotypes differ in gene expression in either tissue, (ii) which biological
processes distinguish them, and (iii) whether the two tissues carry a shared
signature. We find that the gonad, not the liver, carries the phenotype.

# Materials and methods

## Animals and sampling

**[[To be completed by the authors: collection site(s) and dates; how fish were
identified as territorial or social, including the behavioural criteria and
observation period; sex determination and maturity staging at dissection; body size;
permits and animal-care approval.]]** Thirty adult sockeye salmon were sampled, 15
classified as territorial and 15 as social. Liver, gonad, brain and blood were
dissected from each individual, preserved in RNAlater, and stored at −80 °C on return
to the laboratory. Only liver and gonad were sequenced; brain RNA yields were
insufficient.

## RNA isolation

Approximately 20 mg of tissue was homogenised in 1 ml RNAzol RT (Molecular Research
Center, Cincinnati, OH) and total RNA was purified with the Direct-zol RNA Microprep
kit (Zymo Research, Irvine, CA) including the on-column DNase I step. RNA was
quantified by fluorometry (Qubit 3.0, Life Technologies) and integrity was checked on
a subset of samples from each extraction batch on an Agilent 2100 Bioanalyzer; all
checked samples had RNA integrity numbers above 7.

## Library preparation and sequencing

Libraries were prepared and sequenced at the Genomic Sequencing and Analysis
Facility, University of Texas at Austin, using the QuantSeq 3′ mRNA-Seq FWD protocol
(Lexogen, Vienna, Austria), a commercial implementation of TagSeq [@Moll2014;
@Lohman2016]. The method generates a single cDNA fragment from the 3′ end of each
polyadenylated transcript, so tag counts are direct proxies for transcript abundance
and are independent of transcript length [@Meyer2011]. Single-end reads were
sequenced **[[platform and read length to be confirmed]]** in two submissions
(gonad, GSAF job JA22192; liver, JA22330).

## Read processing and quantification

Reads were trimmed with cutadapt [@Martin2011]: the first 15 bases were removed to
eliminate the random priming barcode, poly-A and poly-G runs of eight or more bases
and the Illumina adapter (AGATCGG) were trimmed, and reads shorter than 20 bases were
discarded. Quality was assessed before and after trimming with FastQC and MultiQC
[@Ewels2016]. Trimmed reads were aligned to the sockeye salmon reference assembly
GCF_006149115.2 (Oner_1.1) [@Christensen2020] with HISAT2 v2.2.1 [@Kim2019] using
`--dta` and a relaxed score minimum (`--score-min L,0,-0.6`), and alignments were
sorted with samtools v1.12. Transcripts were assembled per sample against the RefSeq
mRNA annotation with StringTie v2.2.1 [@Pertea2015; @Kovaka2019], merged per tissue,
and gene-level counts were extracted with prepDE.py. Mean overall alignment rates
were 88.7% (SD 2.2) for gonad and 86.5% (SD 0.8) for liver libraries. Two gonad
samples (C05 and C17) clustered apart from all others in the sample-correlation
heatmap and were excluded before modelling, leaving 28 gonad and 30 liver libraries.

## Differential expression

Gene counts were analysed in R 4.3.2 with DESeq2 v1.42.1 [@Love2014] separately for
each tissue, with behavioural phenotype as the single factor and *social* as the
reference level, so that positive fold changes denote higher expression in
territorial fish. Samples were matched between the count matrix and the phenotype
table by identifier and the match was asserted programmatically. Dispersions and
size factors were estimated on the full matrix of `r fmt(n_genes)` annotated genes;
for reporting, genes were retained if they had at least 10 counts in at least a third
of the samples (`r fmt(l_kept)` liver and `r fmt(g_kept)` gonad genes). The Wald test
p-values were adjusted for multiple testing with the Benjamini–Hochberg procedure
[@Benjamini1995] after DESeq2's independent filtering, with the filtering target set
to the same 0.05 used for significance. Log~2~ fold changes were then shrunk with the
apeglm estimator [@Zhu2019]; shrinkage was applied to that results table, so it
changes fold-change estimates but not p-values or significance calls. Genes with an
adjusted p below 0.05 were considered differentially expressed. Ordination used principal component analysis of
variance-stabilised counts; sample clustering used Pearson correlation of
regularised-log counts.

## Functional enrichment

Gene Ontology (GO) annotations were taken from the NCBI annotation release for this
assembly (GAF 2.2, 139,236 distinct gene–term pairs after removing NOT-qualified
annotations, covering 32,629 genes and 6,937 terms), and KEGG pathway membership from
the KEGG REST service for organism `one` (12,285 genes in 196 pathways); both were
vendored into the repository with their retrieval dates. For each tissue the test
universe was every gene that received an adjusted p-value, and the differentially
expressed set was those with adjusted p below 0.05, the same rule used for the
results tables. Over-representation was tested with the hypergeometric test as
implemented in goseq [@Young2010] and corrected across tested categories by
Benjamini–Hochberg. We did not use goseq's default Wallenius test, which re-weights
genes by transcript length, for two reasons. First, TagSeq counts one fragment per
transcript, so the length-dependent power that the weighting is designed to remove is
absent by construction. Second, the relationship between length and differential
expression in these data runs the other way: in the gonad, `r formatC(dec1, format = "f", digits = 0)`%
of genes in the shortest length decile were differentially expressed against
`r formatC(dec_rest[1], format = "f", digits = 0)`–`r formatC(dec_rest[2], format = "f", digits = 0)`% in every
other decile. The Wallenius weighting is estimated by fitting a monotone spline to
that relationship, and with an inverted relationship the fit proved numerically
unstable: across computing platforms and near-identical gene sets it either
suppressed the short-gene pathways that constitute the biological signal or
collapsed to a constant. The hypergeometric test is exact and deterministic.

## Reproducibility

Every result reported here is produced by rendering the Quarto notebooks in the
project repository, with R package versions pinned in a renv lockfile against a dated
CRAN snapshot and Bioconductor 3.18. Input files are checksummed and verified at the
start of each render. A GitHub Actions workflow restores the environment on a clean
Ubuntu runner, re-renders the analysis end to end, and fails if any result table
changes in rows, columns or significance calls at 0.05; the tables reported below have
passed that check. The manuscript itself is a Quarto document whose numbers are read
from the committed result tables at render time.

# Results

## Sequencing and sample quality

Libraries yielded `r depth_txt("gonad")` gene-assigned counts per gonad sample and
`r depth_txt("liver")` per liver sample (mean ± SD). Of the `r fmt(n_genes)`
annotated genes, `r fmt(g_kept)` (gonad) and `r fmt(l_kept)` (liver) passed the
minimum-count filter, and `r fmt(g_tested)` and `r fmt(l_tested)` respectively
received an adjusted p-value after independent filtering. Expression of sex-linked
marker genes was uniform across gonad samples: forkhead box L2 paralogues were highly
expressed in every sample (112–410 counts per million), zona pellucida genes were
expressed in all, and the testis markers *dmrt1* and *amh* were low throughout,
indicating that all 28 gonad libraries were ovarian **[[to be confirmed against
dissection records]]**. Zona pellucida transcript abundance varied more than
100-fold among fish, consistent with heterogeneity in oocyte developmental stage.

## Global expression structure

Neither tissue separated cleanly by phenotype in unsupervised ordination
(@fig-overview A, D). In the liver the first two principal components explained
`r pcv("liver", "PC1")` and `r pcv("liver", "PC2")` of variance and the phenotype groups
overlapped almost completely. In the gonad the first component explained
`r pcv("gonad", "PC1")` of variance and territorial samples
were spread along it, with roughly half forming a cluster at negative PC1 scores
that contained no social fish, while the remainder overlapped the social group. The
gonad sample-correlation matrix showed the same structure: a block of six
territorial fish (C01, C06, C07, C08, C10, C11) correlated more strongly with one
another than with any social sample. Phenotype thus accounts for part, but not the
majority, of transcriptome-wide variation in either tissue.

![Transcriptome-wide comparison of territorial (blue circles) and social (orange triangles) sockeye salmon in liver (A–C) and gonad (D–F). (A, D) Principal component analysis of variance-stabilised counts, with 95% confidence ellipses for each phenotype; axis labels give the variance explained. (B, E) Volcano plots of all tested genes. Genes differentially expressed at adjusted p < 0.05 are coloured by the phenotype in which expression is higher, and the counts in the upper corners give the number in each direction; other genes are grey. Labelled genes are discussed in the text. (C, F) Z-scored regularised-log expression of every differentially expressed gene, with genes and samples ordered by hierarchical clustering; the bar above each heatmap gives the phenotype of each sample.](../figures/fig1-overview.png){#fig-overview}

## Liver

Only `r fmt(l_sig)` of `r fmt(l_tested)` tested liver genes differed between
phenotypes at an adjusted p below 0.05 (@fig-overview B, C; @tbl-liver), and fold
changes were modest (median |log~2~ fold change| `r formatC(l_medlfc, format = "f", digits = 2)`;
one gene beyond two-fold). `r fmt(l_up)` genes were more highly expressed in
territorial fish and `r fmt(l_down)` in social fish. The territorial-high set was
dominated by subunits of the mitochondrial respiratory chain and ATP synthase
(NADH dehydrogenase subunits NDUFS5, NDUFA8 and NDUFB5, cytochrome c1, cytochrome
b-c1 complex subunit 9, ATP synthase F~0~ subunit B1 and ATP5MPL), cytosolic
ribosomal proteins (L3, L32 and S13), proteasome subunits, ferritin heavy and middle
subunits, and a mannose-binding protein C-like lectin. The social-high set included
an alpha-tubulin N-acetyltransferase, a disintegrin and metalloproteinase (ADAM10),
unconventional myosin XVI, long-chain acyl-CoA synthetase 1 and low-density
lipoprotein receptor-related protein 1. No GO term or KEGG pathway was enriched
(`r lgo_sig` and `r lk_sig` at FDR < 0.05); the closest was
`r tolower(lk_top$name)` (`r lk_top$numDEInCat` of `r lk_top$numInCat` pathway genes,
FDR `r formatC(lk_top$FDR, format = "f", digits = 2)`).

```{r}
#| label: tbl-liver
#| tbl-cap: !expr sprintf("All %d liver genes differentially expressed between territorial and social sockeye (adjusted p < 0.05). Positive fold changes indicate higher expression in territorial fish. LOC identifiers are NCBI gene symbols for genes without an assigned name.", l_sig)
kable(gene_tbl(L$sig, by = "lfc"), align = c("l", "l", "r", "r"))
```

## Gonad

The gonad responded strongly: `r fmt(g_sig)` of `r fmt(g_tested)` tested genes were
differentially expressed (@fig-overview E, F), of which `r fmt(g_up)`
(`r pct(g_up, g_sig)`) were more highly expressed in territorial fish and
`r fmt(g_down)` in social fish. The median |log~2~ fold change| was
`r formatC(g_medlfc, format = "f", digits = 2)` and `r fmt(g_lfc1)` genes changed more
than two-fold. The largest effects were increases in territorial fish of genes of
the acute-phase and innate immune response: two serum amyloid A-5 paralogues and an
equistatin-like protease inhibitor were each about 60-fold higher (log~2~ fold change
6.0–6.2), and ladderlectin-like, saxiphilin-like and neurofibromin-like genes were
30–40-fold higher (@tbl-gonad; @fig-genes A). The most statistically robust changes were an
increase in the small heat shock protein *hspb11* and the centrosomal protein
*cep170*. Genes higher in social fish included vasotocin-neurophysin VT1,
a CD209 antigen-like lectin, dentin sialophosphoprotein-like and an inward-rectifier
potassium channel.

```{r}
#| label: tbl-gonad
#| tbl-cap: "The 20 gonad genes with the largest absolute shrunken fold change between territorial and social sockeye (adjusted p < 0.05). Positive values indicate higher expression in territorial fish."
kable(gene_tbl(G$sig, n = 20, by = "lfc"), align = c("l", "l", "r", "r"))
```

## Functional enrichment in the gonad

Over-representation analysis recovered a coherent programme (@fig-enrichment;
@tbl-kegg, @tbl-go).
`r fmt(k_sig)` KEGG pathways were enriched at FDR < 0.05, led by oxidative
phosphorylation (`r oxphos$numDEInCat` of `r oxphos$numInCat` pathway genes
differentially expressed, FDR `r sci(oxphos$FDR)`) and ribosome
(`r ribo$numDEInCat` of `r ribo$numInCat`, FDR `r sci(ribo$FDR)`), followed by
`r knitr::combine_words(k_rest)`. The cardiac muscle contraction pathway is enriched
here for the same reason as oxidative phosphorylation: most of its differentially
expressed members are subunits of respiratory complexes III and IV. The direction of
these changes was almost entirely one-sided: all `r ox_dir["n"]` differentially
expressed oxidative phosphorylation genes and `r ribo_dir["up"]` of `r ribo_dir["n"]`
ribosomal genes were higher in territorial fish. The `r fmt(go_sig)` enriched GO
terms told the same story from the other annotation: structural constituent of
ribosome, translation, cytosolic large ribosomal subunit, mitochondrial respirasome
and respiratory chain complexes I, III and IV, ATP synthase, spliceosomal snRNPs,
RNA polymerase II core complex, prefoldin and chaperone-mediated complex assembly.
Together these describe gonadal tissue in territorial fish that is synthesising more
protein and generating more ATP than that of social fish.

```{r}
#| label: tbl-kegg
#| tbl-cap: "KEGG pathways over-represented among gonad genes differentially expressed between territorial and social sockeye (hypergeometric test, FDR < 0.05). Universe: all genes with an adjusted p-value that map to at least one pathway."
G$kegg %>% filter(FDR < 0.05) %>%
  transmute(Pathway = category, Name = name, `DE genes` = numDEInCat, `Pathway genes` = numInCat,
            p = sapply(over_represented_pvalue, sci), FDR = sapply(FDR, sci)) %>%
  kable(align = c("l", "l", "r", "r", "r", "r"))
```

```{r}
#| label: tbl-go
#| tbl-cap: !expr sprintf("Gene Ontology terms over-represented among gonad genes differentially expressed between territorial and social sockeye (hypergeometric test, FDR < 0.05; %d terms). BP, biological process; CC, cellular component; MF, molecular function.", go_sig)
G$go %>% filter(FDR < 0.05) %>%
  transmute(Term = category, Description = term, Ontology = ontology, `DE genes` = numDEInCat, `Term genes` = numInCat,
            FDR = sapply(FDR, sci)) %>%
  kable(align = c("l", "l", "c", "r", "r", "r"))
```

![KEGG pathways and Gene Ontology terms over-represented among gonad genes differentially expressed between territorial and social sockeye (hypergeometric test, FDR < 0.05). The horizontal position is the share of each category's genes that are differentially expressed; the dashed line marks the same share across all tested genes. Point size gives the number of differentially expressed genes in the category and colour its FDR. Categories are ordered by that share within each panel.](../figures/fig2-enrichment.png){#fig-enrichment}

## Overlap between tissues

`r length(shared)` genes were differentially expressed in both tissues
(`r paste(shared, collapse = ", ")`), and `r shared_same_dir` of them changed in the
same direction in both (@fig-genes B). They include NADH dehydrogenase subunits NDUFS5 and NDUFB5,
ATP5MPL, a cytochrome b-c1 complex subunit, the oligosaccharyltransferase subunit
OSTC, proteasome subunit SEM1 and splicing factor 3B subunit 6, each higher in
territorial fish in both tissues. This is a small set, but its composition mirrors at
the level of individual genes the translation and energy-metabolism programme that
dominates the gonad.

![Named genes and genes shared between tissues. (A) Normalised counts in gonad for the genes discussed by name in the text, by phenotype (territorial, blue circles; social, orange triangles), on a log scale; boxes show the median and interquartile range. Fold changes and adjusted p-values for all six genes are in the full gonad results table (see Data and code availability). (B) Shrunken log~2~ fold changes in liver and gonad for every gene differentially expressed in both tissues; the dashed line marks equal change in the two tissues, and labelled genes are those named in the text.](../figures/fig3-genes.png){#fig-genes}

# Discussion

Territorial and social sockeye differed sharply in their gonadal transcriptomes and
barely at all in their livers. The gonadal signature has two components. The larger
one, comprising most of the `r fmt(g_sig)` differentially expressed genes and all of
the enriched pathways, is a coordinated increase in the machinery of protein
synthesis and mitochondrial ATP production in territorial fish. The smaller, more
dramatic one is a set of acute-phase and stress-response genes induced ten- to
sixty-fold. We consider each in turn, then the near-silence of the liver, and finally
what this design can and cannot say about cause.

## A gonad working harder

Ribosomal proteins, translation factors, respiratory chain complexes and ATP synthase
are among the most abundant transcripts in any cell, and their coordinated elevation
indicates a tissue with higher biosynthetic and energetic demand rather than a change
in cell identity. In a maturing ovary, that demand is set by oocyte growth: the
accumulation of yolk, the synthesis of zona pellucida and maternal transcripts, and
the mitochondrial proliferation that provisions the egg [@Kleppe2014]. An ovary in
which these processes are more active is an ovary further into, or more heavily
invested in, maturation. The simplest reading of our result is therefore that
territorial females were in a more advanced or more active state of reproductive
investment than social females at the time of capture. This aligns with the
observation across teleosts that dominance and reproductive activation are coupled:
in tilapia, dominant individuals have higher gonadosomatic index, higher circulating
gonadotropins and sex steroids, and testis transcriptomes in which growth-factor,
gonadotropin and steroidogenic pathways are up-regulated [@Thonnes2022;
@HollanderCohen2022], and dominance-associated brain gene expression in cichlids is
enriched for genes with reproductive-state bias [@Renn2016]. In sockeye, a female
holding and defending a nest site is by definition close to spawning, whereas a
female in an aggregation may be staging. The pronounced substructure among
territorial gonads on the first principal component, together with the wide
variation in zona pellucida transcript abundance among fish, suggests that
maturation stage varied within the territorial group as well, and that phenotype is
a coarse proxy for a continuous reproductive state.

The one-sidedness of the pattern deserves emphasis. Every differentially expressed
oxidative phosphorylation gene was higher in territorial fish. In male mice living in
social hierarchies, splenic oxidative phosphorylation genes are instead uniformly
higher in subordinates, and hepatic fatty acid oxidation is higher in dominants
[@Lee2022], while in rainbow trout the same hepatic pathways move in the opposite
direction to mice [@Kostyniuk2018]. Energy-metabolism transcription is evidently
highly responsive to social phenotype but in a tissue- and context-specific way, and
in the salmon ovary the direction is toward greater capacity in the territorial
phenotype.

## Acute-phase and stress genes

The largest fold changes in the data set were induction, in territorial gonads, of two
serum amyloid A paralogues, an equistatin-like cysteine protease inhibitor, a
ladderlectin, a saxiphilin-like transferrin-family protein and the small heat shock
protein *hspb11*. Serum amyloid A is a canonical acute-phase reactant whose
transcription rises by orders of magnitude in response to inflammatory cytokines,
and lectins, protease inhibitors and heat shock proteins are the effectors of the
innate response to tissue damage and infection. Their appearance in ovary rather
than liver suggests a local response within the gonad. Two non-exclusive
explanations fit. Territorial behaviour, especially nest defence, is aggressive and
energetically costly [@Quinn1999; @McVeigh2007] and may carry a systemic
inflammatory and glucocorticoid load that peripheral tissues register [@Gilmour2005;
@SnyderMackler2016]. Alternatively, an ovary approaching ovulation undergoes
follicular remodelling that recruits many of the same innate-immune and
chaperone pathways. Distinguishing these requires the maturity staging and plasma
measurements that the present design lacks.

Among genes higher in social fish, vasotocin-neurophysin VT1 is notable. Vasotocin,
the teleost homologue of vasopressin, is a nonapeptide whose neural expression is
repeatedly associated with dominance, aggression and territoriality in fishes
[@Godwin2012], and it was up-regulated in the brains of dominant cichlid females
[@Renn2016]. Its expression in gonad, where it was more than four-fold lower in
territorial fish, points to peripheral nonapeptide signalling in the ovary that is
patterned by behavioural phenotype, a possibility not previously reported in salmon
and worth targeted follow-up.

## A liver that does not tell the two phenotypes apart

Only `r fmt(l_sig)` hepatic genes differed between phenotypes, with small fold
changes and no pathway enrichment. This contrasts with the marked hepatic metabolic
reorganisation seen in juvenile rainbow trout after a few days of paired
confinement [@Kostyniuk2018] and in group-housed mice [@Lee2022]. Several
differences in design may explain the contrast. Those experiments imposed dyadic or
group hierarchies in the laboratory and sampled during active feeding, when the
liver's role in partitioning ingested energy between growth and storage makes it a
sensitive readout of status. Adult sockeye on the spawning migration do not feed,
and both phenotypes were drawing on the same finite somatic reserves; the liver's
transcriptional programme may be dominated by that shared state. The genes that did
differ are nonetheless interpretable: higher respiratory-chain, ATP-synthase,
ribosomal and ferritin subunits in territorial fish, and lower LRP1 and long-chain
acyl-CoA synthetase 1, are consistent with mild differences in oxidative metabolism,
iron handling and lipid uptake rather than a distinct hepatic phenotype. It is also
possible that the liver responds on a slower timescale than the gonad, or that the
field classification captured a behavioural state too transient to be recorded in
hepatic transcription. The `r length(shared)` genes shared between tissues, which
include the same respiratory-chain and translation-associated genes that dominate
the gonad and which changed in the same direction in both, together with a
non-significant tendency toward enrichment of `r tolower(lk_top$name)` in the liver,
hint that a weak systemic version of the gonadal programme reaches the liver.

## Methodological note on enrichment testing with TagSeq

We report enrichment from the hypergeometric test rather than the length-weighted
Wallenius test that is goseq's default. The default is appropriate for
whole-transcript RNA-seq, where longer transcripts accrue more reads and are called
differentially expressed more readily [@Young2010]. TagSeq removes that dependence
by design [@Meyer2011; @Lohman2016]. In our gonad data the length association was
strong but inverted: short genes were five times more likely to be differentially
expressed than long ones, because ribosomal and respiratory-chain genes are short
and were the genes that changed. Where the Wallenius weighting fitted, it therefore
removed the signal rather than a bias: in an earlier version of this analysis that
differed only in the independent-filtering target, it reduced the enriched KEGG
pathways from six to one on one operating system and to zero on another. On the
final gene set the weighting spline could not be fitted at all and collapsed to a
constant, returning the hypergeometric result. A correction whose output ranges from
none to all of the enriched pathways across near-identical inputs is not usable. We
flag this because length-bias correction is often applied to 3′-tag data by default,
and in data sets where the differentially expressed genes are themselves short it
will either hide the result or fail without warning.

## Limitations

The design is observational and cross-sectional. Fish were classified by behaviour
at capture and sampled once, so we cannot distinguish a gonadal state that causes
territorial behaviour from one that accompanies or follows it, and we cannot exclude
that phenotype assignment partly reflects maturation stage. Sex and maturity were not
part of the metadata available to this analysis; we infer from marker-gene
expression that all gonads were ovarian, but this must be verified against
dissection records, and the analysis should be repeated with maturity stage as a
covariate if it was recorded. Two gonad samples were excluded on the basis of a
correlation heatmap without a pre-registered criterion. The sample size, 15 per
group, gives good power for the large gonadal effects but limited power in the
liver, where the true number of affected genes is probably larger than
`r fmt(l_sig)`.

# Conclusions

In adult sockeye salmon, the territorial phenotype is written in the gonad. Territorial
females carried an ovarian transcriptome with elevated translation and oxidative
phosphorylation and strong induction of acute-phase genes, whereas their livers were
almost indistinguishable from those of social females. The results identify
reproductive investment, rather than hepatic energy metabolism, as the physiological
axis along which behavioural type varies in spawning-phase sockeye, and they offer a
set of candidate genes, from serum amyloid A to vasotocin, for testing whether
gonadal state drives behaviour or behaviour drives gonadal state.

# Data and code availability

Count matrices, phenotype tables, annotation resources, all result tables and the
complete analysis are in the project repository at
<https://github.com/RobertsLab/project-sockeye-tagseq>, and the rendered analysis is
published at <https://robertslab.github.io/project-sockeye-tagseq/>. Raw FASTQ files
are currently hosted at
<https://gannet.fish.washington.edu/panopea/berdahl-sockeye-salmon/> and will be
deposited in the NCBI Sequence Read Archive **[[BioProject accession to be added]]**.

# Author contributions and funding

**[[To be completed.]]**

# References

::: {#refs}
:::
