---
title: "Differential expression"
subtitle: "Territorial vs social sockeye salmon"
date: today
params:
  tissue: liver
execute:
  warning: true
  message: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
```

This notebook runs the DESeq2 analysis for **one tissue**, chosen by the
`tissue` parameter: territorial against social fish, with social as the
reference, reporting genes at adjusted p < 0.05.

`./render-all.sh` renders it for both tissues. To render one:

```bash
quarto render tag-seq/code/02-differential-expression.qmd -P tissue:liver
quarto render tag-seq/code/02-differential-expression.qmd -P tissue:gonad
```

Package versions are pinned in `renv.lock`; run `renv::restore()` once on a
fresh clone. Nothing is installed at render time.

## Setup

```{r}
#| label: load_libraries
library(DESeq2)       # DESeqDataSetFromMatrix, DESeq, results, lfcShrink, vst, rlog
library(apeglm)       # lfcShrink(type = "apeglm")
library(ashr)         # lfcShrink(type = "ashr")
library(PCAtools)     # pca, screeplot, biplot, pairsplot, getComponents
library(pheatmap)     # sample-correlation heatmap
library(RColorBrewer) # brewer.pal, for the DEG heatmap palette
library(ggpubr)       # ggexport
library(ggplot2)
library(dplyr)
library(tibble)       # rownames_to_column

source("tag-seq/code/_common.R")

cfg <- tissue_config(params$tissue)
dir.create(cfg$dir, recursive = TRUE, showWarnings = FALSE)
cfg[c("tissue", "prefix", "dir", "exclude")]
```

## Load data

Samples are matched between the count matrix and the phenotype table **by
name**, not by column position. `prepDE.py` names columns after the GTF files
(`01B_S43_R1.gtf`); the phenotype tables use `B01`. `load_counts()` derives one
from the other and asserts the two sets are identical before aligning them.
Gene identifiers are the part of each StringTie row name after the pipe
(`gene-arhgap8|arhgap8` gives `arhgap8`, `gene-LOC115144855|LOC115144855` gives
`LOC115144855`), which is unique across all 37,942 genes, and descriptions come
from the NCBI feature table for the assembly, which covers 37,929 of them.

```{r}
#| label: load_data
annotation    <- load_gene_annotation()
feature_table <- annotation[, c("gene", "description")]

dat     <- load_counts(cfg)
cts     <- dat$cts
coldata <- dat$coldata

print(dplyr::count(coldata, phenotype))
cat("samples:", ncol(cts), " genes:", nrow(cts), "\n")
cat("reference level:", levels(coldata$phenotype)[1], "\n")
```

## Fit the model

The model is fitted by `fit_deseq()` in `_common.R`, which notebook 05 also
calls to draw the manuscript figures, so the tables and the figures come from
one definition of the model. It does three things:

1. Fits `~ phenotype` with DESeq2.
2. Asserts that coefficient 2 is `phenotype_territorial_vs_social`, which
   holds because `load_counts()` sets `social` as the reference level.
3. Keeps genes with 10 or more counts in at least a third of samples. This
   filtering happens *after* `DESeq()`, so dispersions and size factors are
   estimated on the full matrix.

```{r}
#| label: deseq
fit      <- fit_deseq(dat)
dds      <- fit$dds
n_before <- fit$n_before
resultsNames(dds)
cat("genes before filtering:", n_before, "\n")
cat("genes after filtering: ", nrow(dds), "\n")
```

## Ordination and sample clustering

```{r}
#| label: pca
#| fig-width: 8
#| fig-height: 7
p <- pca(assay(vst(dds)), metadata = coldata)

screeplot(p, axisLabSize = 18, titleLabSize = 22)

bp1 <- biplot(p, x = "PC1", y = "PC2",
              colby    = "phenotype",
              colkey   = PHENOTYPE_COLOURS,
              shape    = "phenotype",
              shapekey = c(territorial = 19, social = 17),
              pointSize = 4,
              showLoadings = FALSE,
              lab = NULL,
              drawConnectors = FALSE,
              ellipse = TRUE, ellipseLevel = 0.95, ellipseFill = FALSE,
              ellipseAlpha = 1/4, ellipseLineSize = 1,
              xlim = cfg$pca_xlim, ylim = c(-100, 100),
              gridlines.major = FALSE, gridlines.minor = FALSE,
              borderWidth = 1,
              legendPosition = "top", legendLabSize = 16, legendIconSize = 6.0) +
  theme(axis.text.x  = element_text(size = 16, colour = "black"),
        axis.text.y  = element_text(size = 16, colour = "black"),
        axis.ticks.x = element_line(colour = "black"),
        axis.ticks.y = element_line(colour = "black"))
bp1

bp2 <- pairsplot(p,
                 components = getComponents(p, c(1:5)),
                 triangle = TRUE, trianglelabSize = 12,
                 hline = 0, vline = 0,
                 pointSize = 1,
                 gridlines.major = FALSE, gridlines.minor = FALSE,
                 colby = "phenotype",
                 title = "Pairs plot", plotaxes = FALSE,
                 margingaps = unit(c(-0.01, -0.01, -0.01, -0.01), "cm"))
bp2

rld     <- rlog(dds, blind = TRUE)
rld_cor <- cor(assay(rld))
bp3     <- pheatmap(rld_cor, annotation = coldata[, 1:2])
bp3
```

The ordination and clustering figures are written to this tissue's results
directory.

```{r}
#| label: export_ordination
ggexport(filename = out_path(cfg, cfg$prefix, "-PCA.png"),
         plot = bp1, res = 600, width = 4000, height = 5000)

## The pairs plot of PCs 1-5 is a diagnostic, kept out of git (alternatives/).
ggexport(filename = alt_path(cfg, cfg$prefix, "-PAIRS.png"),
         plot = bp2, res = 600, width = 6000, height = 6000)

ggexport(filename = out_path(cfg, cfg$prefix, "-pheatmap.png"),
         plot = bp3, res = 600, width = 6000, height = 6000)
```

## Contrasts and shrinkage

The results are computed once, with `alpha = 0.05`, and that object is passed
to every shrinkage estimator through `res =`. Shrinkage changes only the fold
change and its standard error; the p-values, the adjusted p-values and the set
of genes that survive independent filtering all come from `res_table`, so every
table below makes the same significance calls.

`alpha` matters here because it is the target DESeq2's independent filtering
optimises for. Called without `res =`, `lfcShrink()` computes its own results
with the default `alpha = 0.1`, which filters a different set of low-count
genes and so produces different adjusted p-values; the shrunken tables would
then disagree with the unshrunken one about which genes are significant.

```{r}
#| label: contrasts
res_table        <- results(dds, contrast = c("phenotype", "territorial", "social"), alpha = 0.05)
res_table_normal <- lfcShrink(dds, coef = 2, res = res_table, type = "normal")
res_table_apeglm <- lfcShrink(dds, coef = 2, res = res_table, type = "apeglm")
res_table_ashr   <- lfcShrink(dds, coef = 2, res = res_table, type = "ashr")

## Every table must carry the same adjusted p-values as res_table.
for (shrunk in list(res_table_normal, res_table_apeglm, res_table_ashr)) {
  stopifnot(identical(rownames(shrunk), rownames(res_table)),
            identical(shrunk$padj, res_table$padj))
}
```

Three shrinkage estimators are compared. They move fold changes by different
amounts (the MA plots below), but because they share the unshrunken test's
adjusted p-values they flag exactly the same genes, so the choice affects
effect sizes, never which genes are called. The manuscript reports apeglm. Its
tables are tracked in git; the normal and ashr tables are regenerated on every
render into `alternatives/`, which git ignores, and their fold changes are
published in Tables S2 and S3.

```{r}
#| label: ma_plots
#| fig-width: 10
#| fig-height: 4
draw_ma <- function() {
  par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
  xlim <- c(1, 1e5); ylim <- c(-3, 3)
  for (nm in c("normal", "apeglm", "ashr")) {
    DESeq2::plotMA(get(paste0("res_table_", nm)), xlim = xlim, ylim = ylim, main = nm, cex = .8)
    abline(h = c(-1.5, 1.5), col = "dodgerblue", lwd = 2)
  }
}
draw_ma()

## PNG: as a vector PDF, tens of thousands of points made 2-3 MB per tissue.
png(filename = out_path(cfg, "MA_plots.png"), width = 10, height = 4, units = "in", res = 200)
draw_ma()
invisible(dev.off())
```

## Results tables

Significance is `padj < 0.05`, and only that: no fold-change or s-value
threshold is applied.

```{r}
#| label: tidy_tables
padj.cutoff <- 0.05

as_gene_tibble <- function(x) {
  x %>% data.frame() %>% rownames_to_column(var = "gene") %>% as_tibble()
}

res_table        <- as_gene_tibble(res_table)
res_table_normal <- as_gene_tibble(res_table_normal)
res_table_apeglm <- as_gene_tibble(res_table_apeglm)
res_table_ashr   <- as_gene_tibble(res_table_ashr)

cat("Total genes:", nrow(res_table), "\n")
cat("Genes after NAs removed:", nrow(na.omit(res_table)), "\n")

res_table        <- na.omit(res_table)
res_table_normal <- na.omit(res_table_normal)
res_table_apeglm <- na.omit(res_table_apeglm)
res_table_ashr   <- na.omit(res_table_ashr)
```

```{r}
#| label: annotate
annotate_genes <- function(tbl) {
  tbl %>%
    left_join(feature_table, by = "gene") %>%
    arrange(-log2FoldChange) %>%
    mutate(tissue = cfg$tissue)
}

all_genes        <- annotate_genes(res_table)
all_genes_normal <- annotate_genes(res_table_normal)
all_genes_apeglm <- annotate_genes(res_table_apeglm)
all_genes_ashr   <- annotate_genes(res_table_ashr)

sig_genes        <- annotate_genes(filter(res_table,        padj < padj.cutoff))
sig_genes_normal <- annotate_genes(filter(res_table_normal, padj < padj.cutoff))
sig_genes_apeglm <- annotate_genes(filter(res_table_apeglm, padj < padj.cutoff))
sig_genes_ashr   <- annotate_genes(filter(res_table_ashr,   padj < padj.cutoff))
```

The four estimators flag exactly the same genes, because they share one set of
adjusted p-values. That is asserted here, and it is why one table of normalised
counts serves all four: the counts depend on the gene, not on the estimator.

```{r}
#| label: single_gene_counts
stopifnot(setequal(sig_genes$gene, sig_genes_normal$gene),
          setequal(sig_genes$gene, sig_genes_apeglm$gene),
          setequal(sig_genes$gene, sig_genes_ashr$gene))

## Normalised counts for every significant gene.
single_gene_counts <- function(sig) {
  if (nrow(sig) == 0) return(data.frame())
  do.call(rbind, lapply(sig$gene, function(locid) {
    d <- plotCounts(dds, gene = locid, intgroup = "phenotype", returnData = TRUE)
    d$gene   <- locid
    d$sample <- rownames(d)
    rownames(d) <- NULL
    d
  }))
}

write.table(single_gene_counts(sig_genes_apeglm),
            alt_path(cfg, cfg$prefix, "-single-gene-counts.csv"), row.names = FALSE, sep = ",")
```

The summary table `<tissue>-gene-counts.csv` records, per estimator, how many
genes received an adjusted p-value (`genes_tested_*`) and how many of those
have `padj < 0.05` (`significant_*`), together with the gene counts before and
after the reporting filter.

```{r}
#| label: write_tables
gene_counts <- data.frame(count = c(
  nrow(cts), nrow(dds), nrow(cts) - nrow(dds),
  nrow(all_genes), nrow(all_genes_normal), nrow(all_genes_apeglm), nrow(all_genes_ashr),
  nrow(sig_genes), nrow(sig_genes_normal), nrow(sig_genes_apeglm), nrow(sig_genes_ashr)
), row.names = c(
  "genes_before_filtering", "genes_after_filtering", "genes_dropped",
  "genes_tested_unshrunken", "genes_tested_normal", "genes_tested_apeglm", "genes_tested_ashr",
  "significant_unshrunken", "significant_normal", "significant_apeglm", "significant_ashr"
))
gene_counts

## Tracked: the unshrunken test and the apeglm estimates the manuscript uses.
write.table(all_genes,        out_path(cfg, cfg$prefix, "-ALL-DEG.csv"),            row.names = FALSE, sep = ",")
write.table(all_genes_apeglm, out_path(cfg, cfg$prefix, "-ALL-DEG-apeglm.csv"),     row.names = FALSE, sep = ",")
write.table(sig_genes,        out_path(cfg, cfg$prefix, "-SIG-DEG-unshrunken.csv"), row.names = FALSE, sep = ",")
write.table(sig_genes_apeglm, out_path(cfg, cfg$prefix, "-SIG-DEG-apeglm.csv"),     row.names = FALSE, sep = ",")
write.table(gene_counts,      out_path(cfg, cfg$prefix, "-gene-counts.csv"),        row.names = TRUE, sep = ",")

## Regenerated but untracked (alternatives/): the normal and ashr estimates.
## Their fold changes reach the paper through Tables S2 and S3 (notebook 06).
write.table(all_genes_normal, alt_path(cfg, cfg$prefix, "-ALL-DEG-normal.csv"),     row.names = FALSE, sep = ",")
write.table(all_genes_ashr,   alt_path(cfg, cfg$prefix, "-ALL-DEG-ashr.csv"),       row.names = FALSE, sep = ",")
write.table(sig_genes_normal, alt_path(cfg, cfg$prefix, "-SIG-DEG-normal.csv"),     row.names = FALSE, sep = ",")
write.table(sig_genes_ashr,   alt_path(cfg, cfg$prefix, "-SIG-DEG-ashr.csv"),       row.names = FALSE, sep = ",")
```

## Volcano plots

```{r}
#| label: volcanoes
#| fig-width: 6
#| fig-height: 4
volcano <- function(tbl) {
  ggplot(tbl, aes(x = log2FoldChange, y = -log10(pvalue))) +
    geom_point(aes(color = log2FoldChange > 0))
}

## Unshrunken and apeglm, shown here; only apeglm, the estimate the manuscript
## uses, is saved. The manuscript's volcano plots are drawn by notebook 05.
print(volcano(all_genes) + ggtitle("unshrunken"))

p <- volcano(all_genes_apeglm) + ggtitle("apeglm")
print(p)
ggsave(out_path(cfg, "Volcano_all_genes_apeglm.png"), plot = p,
       dpi = 600, device = "png", width = 6, height = 4, units = "in")
```

## Expression heatmap

Z-scored regularised-log expression of every significant gene. The column
colours are looked up per sample from `coldata`, so they follow the samples
whatever the group sizes or column order.

```{r}
#| label: heatmap
sigs <- res_table_apeglm %>%
  filter(padj < 0.05) %>%
  left_join(feature_table, by = "gene")

df_sigs <- as.data.frame(sigs)
stopifnot(!anyDuplicated(df_sigs$gene))
rownames(df_sigs) <- df_sigs$gene

rlog_out <- rlog(dds, blind = FALSE)
mat_sigs <- assay(rlog_out)[rownames(df_sigs), rownames(coldata)]
colnames(mat_sigs) <- rownames(coldata)

## Z-score each gene across samples.
mat_scaled_sigs <- t(apply(mat_sigs, 1, scale))
colnames(mat_scaled_sigs) <- colnames(mat_sigs)

side_colours <- phenotype_side_colours(coldata)
stopifnot(length(side_colours) == ncol(mat_scaled_sigs))

png(filename = out_path(cfg, cfg$prefix, "_heatmap_pval_genes_apeglm.png"),
    res = 1200, width = 6000, height = 6000, units = "px", pointsize = 6)
heatmap(mat_scaled_sigs, scale = "none",
        col = colorRampPalette(brewer.pal(11, "RdYlBu"))(256),
        ColSideColors = side_colours)
invisible(dev.off())
```

## Session

```{r}
#| label: session
sessionInfo()
```
