---
title: "Gene tables"
subtitle: "Splitting the significant DEGs by annotation status"
date: today
---

Splits each tissue's significant genes (apeglm, adjusted p < 0.05) by
annotation status:

1. Genes with **no description** in the NCBI feature table are set aside.
2. Of the rest, **uncharacterized** are those whose description begins with
   `uncharacterized`; everything else is **characterized**.

It writes the two sets per tissue, the list of annotated gene identifiers, and
a summary of the uncharacterized share.

```{r}
#| label: setup
library(dplyr)
source("tag-seq/code/_common.R")

GENE_TABLE_DIR <- "tag-seq/gene_tables"
dir.create(GENE_TABLE_DIR, recursive = TRUE, showWarnings = FALSE)
```

```{r}
#| label: split
split_by_annotation <- function(tissue) {
  cfg <- tissue_config(tissue)
  sig <- read.csv(out_path(cfg, cfg$prefix, "-SIG-DEG-apeglm.csv"),
                  stringsAsFactors = FALSE)

  unannotated <- sig %>% filter(is.na(description))
  annotated   <- sig %>% filter(!is.na(description))
  uncharacterized <- annotated %>% filter(grepl("^uncharacterized", description))
  characterized   <- annotated %>% filter(!grepl("^uncharacterized", description))

  list(tissue = tissue, sig = sig, unannotated = unannotated,
       annotated = annotated, uncharacterized = uncharacterized,
       characterized = characterized)
}

tables <- lapply(c("liver", "gonad"), split_by_annotation)
names(tables) <- c("liver", "gonad")

for (t in tables) {
  cat(sprintf("%-6s significant %5d | no description %4d | characterized %5d | uncharacterized %3d\n",
              t$tissue, nrow(t$sig), nrow(t$unannotated),
              nrow(t$characterized), nrow(t$uncharacterized)))
}
```

The tables are written as quoted CSV, because descriptions contain commas
(`uncharacterized LOC115129809, transcript variant X3`).

```{r}
#| label: write_tables
for (t in tables) {
  write.csv(t$characterized,
            file.path(GENE_TABLE_DIR, paste0(t$tissue, "_characterized.txt")),
            row.names = FALSE)
  write.csv(t$uncharacterized,
            file.path(GENE_TABLE_DIR, paste0(t$tissue, "_uncharacterized.txt")),
            row.names = FALSE)
  writeLines(t$annotated$gene,
             file.path(GENE_TABLE_DIR, paste0(t$tissue, "_all_LOCID.txt")))
}

uncharacterized_stats <- data.frame(
  known           = vapply(tables, function(t) nrow(t$characterized),   integer(1)),
  unknown         = vapply(tables, function(t) nrow(t$uncharacterized), integer(1)),
  row.names       = names(tables)
)
uncharacterized_stats$percent_unknown <-
  uncharacterized_stats$unknown /
  (uncharacterized_stats$unknown + uncharacterized_stats$known)

uncharacterized_stats

write.csv(uncharacterized_stats,
          file.path(GENE_TABLE_DIR, "uncharacterized_stats.txt"),
          row.names = TRUE)
```

```{r}
#| label: session
sessionInfo()
```
