---
title: "Supplementary tables"
subtitle: "Tables S1–S6, assembled from the committed result tables"
date: today
execute:
  warning: true
  message: false
---

This notebook writes the manuscript's supporting-information tables to
`manuscript/supplementary/`, one CSV per table. It computes no new statistics:
every value comes from the count matrices or from the tables notebooks 02 and
04 have already written, and the joins below assert that those tables agree
with one another before anything is written.

The tables are CSV rather than Excel: CSV is the format data archives prefer,
and writing Excel would add a package to `renv.lock` for no gain in content.
Column definitions are in `manuscript/supplementary/README.md`, and the table
legends are in the manuscript's Supporting information section.

| Table | Contents |
|---|---|
| S1 | Samples: tissue, phenotype, inclusion, library size and genes detected |
| S2 | Liver differential expression, every tested gene, all shrinkage estimators |
| S3 | Gonad differential expression, same layout |
| S4 | GO over-representation, every tested term, both tissues |
| S5 | KEGG over-representation, every tested pathway, both tissues |
| S6 | Genes differentially expressed in both tissues |

```{r}
#| label: setup
library(dplyr)
source("tag-seq/code/_common.R")

SUPP_DIR <- "manuscript/supplementary"
dir.create(SUPP_DIR, recursive = TRUE, showWarnings = FALSE)

write_supp <- function(x, name) {
  path <- file.path(SUPP_DIR, paste0(name, ".csv"))
  write.csv(x, path, row.names = FALSE, na = "")
  cat(sprintf("%-40s %6d rows x %2d columns\n", path, nrow(x), ncol(x)))
  invisible(path)
}
```

## Table S1 — samples

Library size is the total of gene-assigned counts in the count matrix, and
genes detected counts genes with at least one read, and with at least ten.
The two excluded gonad libraries are listed with their values, because the
reader should be able to see what was removed.

Sex, maturity stage, body size and per-sample alignment rate are not included
here yet: they are not in the repository (submission checklist A1, A2 and A5).

```{r}
#| label: s1
sample_table <- function(tissue) {
  cfg <- tissue_config(tissue)
  excluded <- cfg$exclude
  cfg$exclude <- character(0)            # load every library, then flag
  dat <- load_counts(cfg)
  data.frame(
    sample             = colnames(dat$cts),
    tissue             = tissue,
    phenotype          = as.character(dat$coldata$trt),
    included           = !colnames(dat$cts) %in% excluded,
    exclusion_reason   = ifelse(colnames(dat$cts) %in% excluded,
                                "outlier in sample-correlation clustering", ""),
    library_size       = colSums(dat$cts),
    genes_detected     = colSums(dat$cts >= 1),
    genes_detected_10  = colSums(dat$cts >= 10),
    row.names = NULL
  )
}
s1 <- bind_rows(sample_table("liver"), sample_table("gonad"))
stopifnot(sum(!s1$included) == 2,
          all(table(s1$tissue, s1$phenotype) == 15))
write_supp(s1, "Table_S1_samples")

## How the excluded libraries compare with the rest of their tissue.
s1 %>% group_by(tissue, included) %>%
  summarise(n = n(), median_library = median(library_size),
            min_library = min(library_size), .groups = "drop")
s1 %>% filter(!included)
```

## Tables S2 and S3 — differential expression

One row per tested gene. The unshrunken Wald test supplies the p-values and
adjusted p-values; each shrinkage estimator supplies its own fold change and
standard error. Since the shrinkage calls reuse the unshrunken results
(notebook 02), the adjusted p-values must be identical across all four source
tables, and that is asserted here.

```{r}
#| label: s2-s3
de_table <- function(tissue) {
  cfg <- tissue_config(tissue)
  rd  <- function(suffix) read.csv(out_path(cfg, cfg$prefix, "-ALL-DEG", suffix, ".csv"),
                                   stringsAsFactors = FALSE)
  un <- rd(""); ap <- rd("-apeglm"); as <- rd("-ashr"); no <- rd("-normal")

  for (x in list(ap, as, no)) {
    stopifnot(setequal(x$gene, un$gene))
    x <- x[match(un$gene, x$gene), ]
    stopifnot(identical(x$padj, un$padj), identical(x$pvalue, un$pvalue))
  }
  pick <- function(x, tag) {
    x <- x[match(un$gene, x$gene), c("log2FoldChange", "lfcSE")]
    setNames(x, paste0(c("log2FC_", "lfcSE_"), tag))
  }

  bind_cols(
    un %>% transmute(gene, description, baseMean,
                     log2FC_unshrunken = log2FoldChange, lfcSE_unshrunken = lfcSE,
                     wald_stat = stat, pvalue, padj),
    pick(ap, "apeglm"), pick(as, "ashr"), pick(no, "normal")
  ) %>%
    mutate(significant = padj < 0.05) %>%
    arrange(padj, pvalue)
}

s2 <- de_table("liver")
s3 <- de_table("gonad")

## Must agree with the significant-gene tables every other notebook reads.
for (x in list(list(t = s2, cfg = tissue_config("liver")),
               list(t = s3, cfg = tissue_config("gonad")))) {
  sig <- read.csv(out_path(x$cfg, x$cfg$prefix, "-SIG-DEG-apeglm.csv"))
  stopifnot(setequal(x$t$gene[x$t$significant], sig$gene))
}
write_supp(s2, "Table_S2_liver_differential_expression")
write_supp(s3, "Table_S3_gonad_differential_expression")
```

## Tables S4 and S5 — enrichment

Every tested category for both tissues, not only the enriched ones, so the
reader can see what was tested and how close the liver came. The hypergeometric
test is one-sided, for over-representation.

```{r}
#| label: s4-s5
enrich_table <- function(kind) {
  bind_rows(lapply(c("liver", "gonad"), function(tissue) {
    x <- read.csv(file.path("tag-seq/GO_output", tissue, sprintf("%s-%s-all.csv", tissue, kind)),
                  stringsAsFactors = FALSE)
    if (kind == "KEGG") x <- rename(x, term = name) %>% mutate(ontology = "KEGG")
    x %>% transmute(tissue, category, term, ontology,
                    de_genes = numDEInCat, category_genes = numInCat,
                    pvalue = over_represented_pvalue, FDR,
                    enriched = FDR < 0.05) %>%
      arrange(pvalue)
  }))
}
s4 <- enrich_table("GO")
s5 <- enrich_table("KEGG") %>% select(-ontology)

enriched <- function(x, t) sum(x$enriched[x$tissue == t])
c(GO_liver = enriched(s4, "liver"), GO_gonad = enriched(s4, "gonad"),
  KEGG_liver = enriched(s5, "liver"), KEGG_gonad = enriched(s5, "gonad"))

write_supp(s4, "Table_S4_GO_enrichment")
write_supp(s5, "Table_S5_KEGG_enrichment")
```

## Table S6 — genes differentially expressed in both tissues

```{r}
#| label: s6
s6 <- inner_join(
  s2 %>% filter(significant) %>% select(gene, description, liver_log2FC = log2FC_apeglm, liver_padj = padj),
  s3 %>% filter(significant) %>% select(gene, gonad_log2FC = log2FC_apeglm, gonad_padj = padj),
  by = "gene"
) %>%
  mutate(same_direction = sign(liver_log2FC) == sign(gonad_log2FC)) %>%
  arrange(desc(gonad_log2FC))
stopifnot(identical(sort(s6$gene),
                    sort(read.csv("figures/source-data/shared-genes.csv")$gene)))
write_supp(s6, "Table_S6_shared_genes")
s6 %>% select(gene, liver_log2FC, gonad_log2FC, same_direction)
```

## Not included

The split of significant genes into characterized and uncharacterized
(`tag-seq/gene_tables/`) is not a separate table: the `description` column of
Tables S2 and S3 carries the same information, and "uncharacterized" genes are
those whose description begins with that word.

```{r}
#| label: session
sessionInfo()
```
